---
title: "data_exploration"
author: "Erica Griffin"
date: "2025-09-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
df <- read
```

First, looking at some of the data to see what the weather distributions are. 

Plot 1: Baseline cort ~ average daytime temp

```{r plot1}
vis1 <- ggplot(initial_vis, aes(x = avgC_capture_day, y = bleed1_final_cort, color = sex)) +
  geom_point(alpha = 0.5) +     
  coord_cartesian(ylim = c(0, 105)) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_manual(
    values = c("Male" = "#1F77B4", "Female" = "#AEC7E8")
  ) +
  theme_minimal() +
  geom_vline(xintercept = 18.5, color = "red", linetype = "dotted", linewidth = 1)
vis1
```

Plot 2: Baseline cort ~ average 3 day temp

Plot 3: Stress-induced cort ~ average daytime temp

Plot 4: Stress-induced cort ~ average 3 day temp

Plot 5: Dex ~ average daytime temp

Plot 6: Dex ~ average 3 day temp 


Next, looking at some covariations of how some of the variables fit together. 

Plot 7: Encounter day of year ~ average dayime temp

Plot 8: Ecounter day of year ~ baseline cort 

Plot 9: Body mass ~ baseline cort 

